---
title: "Paper case study 1: One pathogen system with vaccination (Measles)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{case_study_1_measles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: cs1_measles_citations.bib
link-citations: yes
---

--


Here, we will use the _serosim_  package to generate a longitudinal serological 
survey of a one pathogen system with vaccination and natural infection. The 
overall structure and principles discussed within this case study are also 
applicable to other vaccine preventable pathogens. In this simulation, 
individuals can either be vaccinated against the pathogen, naturally infected 
with the pathogen, a combination of both or neither. We will set up each of the required 
arguments and models for _runserosim_  in the order outlined in the methods section 
of the paper. 

This case study is built around measles and we are interested in 
tracking measles vaccination and infection across a 10 year period. We will 
conduct a longitudinal serological survey with two sampling times where we 
will use an ELISA kit to measure an individual's measles IgG titer. In this example,
biomarker quantity, titer level and antibody level all mean the same thing. 

We will use previously conducted measles serological surveys and any well 
characterized epidemiological and immunological characteristics of measles to 
parameterize our models wherever possible and make reasonable assumptions for 
other required model parameters wherever information is lacking. 

We narrowed our search to measles serological surveys conducted with the 
Enzygnost ELISA IgG test kits and utilized a measles serology review to select a 
few studies of relevance [@Thompson2016-ek]. We did not conduct an extensive 
search of all measles serological surveys in order to parameterize our models as 
it is beyond the scope of this case study. Our aim was to provide a simple 
example of how users can use the literature to inform required _serosim_  inputs. 
Within each section below, we will briefly explain the rationale behind our 
selected model inputs. We caution users to conduct their own research into the 
models and associated parameters which best align with their disease system and 
biomarker test kits. 


Load necessary packages:
```{r message=FALSE, warning=FALSE,eval=TRUE}
## Install and load serosim 
## devtools::install_github("AMenezes97/serosim")
library(serosim)

## Load additional packages required 
library(tidyverse)
library(data.table)
library(ggplot2)
library(patchwork)
library(reshape2)
```


# 1.1 Simulation Settings
We will simulate monthly time steps across a 10 year period. Therefore, we will 
have 120 time steps. Note that these are arbitrary time steps which will need to 
be scaled to the right time resolution to match any time-based parameters used 
later on.
```{r}
## Specify the number of time periods to simulate 
times <- seq(1,120,by=1) 

## Set simulation settings
simulation_settings <- list("t_start"=1,"t_end"=max(times))
```

# 1.2 Population Demography
For this case study, we are not interested in tracking any demography information
but we are interested in tracking individual's birth time. We will 
use the generate_pop_demography function to create the demography tibble needed 
for _runserosim_ .

Note: The _runserosim_  function, called later on, only requires a demography tibble 
with two columns (individuals and times). 
```{r}
## Specify the number of individuals in the simulation 
N<-100

## Generate the population demography tibble
## See help file(?generate_pop_demography) for more information on function arguments.
## age_min is set to 0 which allows births to occur until the last time step
## Let's assume that no individuals are removed from the population and set prob_removal to 0
demography <- generate_pop_demography(N, times, age_min=0, prob_removal=0)

## Examine the generated demography tibble
summary(demography)
```

# 1.3 Exposure to biomarker mapping
Now, we have to set up the exposure IDs and biomarker IDs for our desired 
simulation. Individuals can be seropositive either by measles natural infection 
or by vaccination and we want to track both exposure types separately. 
Here, we will simulate measles natural infection (exposure_ID=measles_ifxn) and 
measles vaccination (exposure_ID=measles_vacc) both of which will boost the same biomarker, IgG antibodies (biomarker_ID=measles_IgG). Since we are only interested 
in tracking measles IgG antibody levels across time, we only need one biomarker 
ID. _runserosim_  requires that exposure_id and biomarker_id are numeric so we will 
use the reformat_biomarker_map function to create a new version of the biomarker 
map. Users can go directly to numeric biomarker_map if they wish.

Note that the 
reformat_biomarker_map function  will number the exposures and biomarkers in 
alphabetical order so that the first exposure event or biomarker that is listed
will not necessarily be labeled as 1.
```{r}
## Create biomarker map
biomarker_map_original <- tibble(exposure_id=c("measles_ifxn","measles_vacc"),biomarker_id=c("measles_IgG","measles_IgG"))
biomarker_map_original

## Reformat biomarker_map for runserosim
biomarker_map <-reformat_biomarker_map(biomarker_map_original)
biomarker_map
```


# 1.4 Force of Exposure and Exposure Model
Now, we need to specify the foe_pars argument which contains the force of exposure
for all exposure_IDs across all time steps and groups. We also specify the 
exposure model which will be called within _runserosim_  later. The exposure model 
will determine the probability that an individual is exposed to a specific 
exposure event.

Since we did not specify different groups within demography, all individuals 
will automatically be assigned group 1. Therefore, we only need 1 row for 
dimension 1 in foe_pars. Dimension 3 of the foe_array must be in the same order 
as the exposure_id within the biomarker map. For example, the force of exposure 
for exposure_id 1 will be inputted within the foe_pars[,,1]. 

We specified the same force of exposure for all time steps within foe_pars for 
simplicity but users will likely have varying numbers to match real world settings. 

```{r, fig.dim = c(4, 3)}
## Create an empty array to store the force of exposure for all exposure types
## Dimension 1: Group
## Dimension 2: Times
## Dimension 3: Exposure ID 
foe_pars <- array(0, dim=c(1,max(times),n_distinct(biomarker_map$exposure_id)))

## Specify the force of exposure for exposure ID 1 which represents measles natural infection
foe_pars[,,1] <- 0.02

## Specify the force of exposure for exposure ID 2 which represents measles vaccination
foe_pars[,,2] <- 0.04

## Specify a simple exposure model which calculates the probability of exposure directly from the force of exposure at that time step
exposure_model<-exposure_model_simple_FOE
## In this selected model, the probability of exposure is 1-exp(-FOE) where FOE is the force of exposure at that time.

## Examine the probability of exposure over time for the specified exposure model
plot_exposure_model(exposure_mode=exposure_model_simple_FOE, times=times, n_groups = 1, n_exposures = 2, foe_pars=foe_pars)
```

# 1.5 Immunity Model
Here, we specify the immunity model which will determine the probability that an 
exposure event is successful in producing an immunological response. Since we have both vaccination and natural 
infection events, we will use immunity_model_vacc_ifxn_biomarker_prot. With this 
immunity model, the probability of successful measles vaccination exposure 
depends on the number of measles vaccines received prior to time t and the 
individual's age at time t while the probability of successful measles infection 
is dependent on the biomarker quantity, in this case antibody titer, at the time of exposure and an individual's total number of measles infections.

Since measles is a fully immunizing pathogen, we have assumed that individuals 
can only be infected once. Therefore, we have disregarded any complexities 
brought on by boosting events from additional measles exposure. This assumption
can be loosened by removing the limit on infection events. 

Within this selected immunity model, individuals can become infected despite 
prior vaccination if their current biomarker quantity is below the threshold of protection as 
a result of waning or incomplete immunity following vaccination. The 
biomarker quantity mediated protection parameters (also known as titer-mediated protection) used within this immunity model are defined within model_pars which will be loaded in section 1.6. 

The Enzygnost ELISA kit reports that the threshold of seropositivity is 200 mIU/ml 
and we have used this titer value to inform our titer-mediated protection curve. 
In our simulation, individuals with titers below 200 mIU/ml will have a higher probability of becoming infected 
when exposed to measles while individuals with titers above 200 mIU/ml are 
more likely to be protected from infection. It is important to note that, seropositivty and titer 
level is not a perfect correlate of protection against infection as it does not 
directly account for neutralizing titers, cellular immunity and other 
immunological complexities. With a more in-depth literature review on correlates 
of protection, we expect users will be able to select better titer-mediated 
protection parameters than our crude estimates. 

```{r, fig.dim = c(4, 4.5)}
## Specify immunity model within runserosim function below 
immunity_model<-immunity_model_vacc_ifxn_biomarker_prot
## This immunity model requires 3 additional arguments specified below. 

## Specify which exposure IDs represent vaccination events 
vacc_exposures<-2

## Specify the age at which an individual is eligible for measles vaccination (9 months old); note non vaccine exposures are listed as NAs
vacc_age<-c(NA,9)

## Specify the maximum number of successful exposure events an individual can receive for each exposure type
## For this example, we are assuming that there is only one dose of the measles vaccine and one possible measles infection
max_events<-c(1,1)

## Plot biomarker-mediated protection curve (also known as titer-mediated protection) given parameters specified within model_pars for biomarker 1 (measles_IgG) which will be loaded in section 1.6
plot_biomarker_mediated_protection(0:1000, biomarker_prot_midpoint=200, biomarker_prot_width=1)
```

# 1.6  Antibody Model, Antibody Kinetics Parameters, and draw_parameters
Now, we specify the antibody model to be used within _runserosim_ to track antibody kinetics, or more broadly biomarker kinetics for each biomarker produced from successful exposure events.

We will be using a biphasic boosting-waning model. This model assumes that for 
each exposure there is a set of long-term boost, long-term boost waning, 
short-term boost, and short-term boost waning parameters. Given varying measles 
antibody dynamics observed in early and late life, we decided that a biphasic 
boosting-waning model was appropriate for our purposes
[@Carryn2019-qv; @Amanna2007-pf].

The antibody kinetics parameters needed for the antibody model are pre-loaded
within a csv file which will take on the argument name: model_pars.
Users can edit the csv file to modify any parameters. _runserosim_  requires that 
exposure_id and biomarker_id are numeric within model_pars so we will use the 
reformat_biomarker_map function again to create a new version of model_pars. 
Users can go directly to numeric model_pars if they wish. 

We selected a few measles serological surveys conducted with the Enzygnost ELISA 
IgG test kits which measured titers following vaccination to inform the vaccine 
induced boost and waning parameters. From the reported quantitative measles IgG 
titers following vaccination, we picked the conservative range of 2500-4500 mIU/mL for vaccine acquired 
full boost. After 10 years, only the long-term boost remains which ranges from 
900-2000 mIU/mL. Therefore, the short-term boost ranges from 1600-2500 mIU/mL
[@Berry2017-ci; @Lyamuya1999-nf; @Carryn2019-qv; @Trevisan2021-of]. Since we are only simulating a 10 year period, we assumed that there was minimal waning of the long-term boost for simplicity.

Although there is evidence that titers following measles infection are 
significantly higher than after vaccination when measured with PRNT assays 
[@Anichini2020-jj], we were unable to find studies that reported titer kinetics 
following known infection events using Enzygnost ELISA IgG test kits. For 
simplicity, we will assume that natural infection boosting parameters are 25% 
higher than vaccine induced antibody boosting when measured with Enzygnost ELISA 
IgG test kits which does not distinguish between neutralizing and non-neutralizing titers.

Lastly, we define the draw_parameters function which determines how each 
individual’s antibody kinetics parameters are simulated from the within-host 
processes parameters tibble (model_pars). We will use a function which draws 
parameters directly from model_pars for the antibody model with random 
effects to represent individual heterogeneity in immunological responses. 
Parameters are drawn randomly from a distribution with mean and standard 
deviation specified within model_pars. This draw_parameters function also 
incorporates biomarker quantitiy dependent boosting (also known as titer-dependent boosting) where an individual’s realized boost is 
dependent on their titer level at the time of the exposure event.

There is contradictory evidence of whether there is a significant boost after 
vaccinated individuals receive another measles vaccine 
[@Christenson1994-fj; @Trevisan2021-of; @Sasaki2019-gl]. For simplicity, we have assumed that 
individuals in this simulation are only eligible for one dose of the measles 
vaccine so we can ignore multiple vaccination events. 

Lastly, it has been reported that there is no significant boost in titers with 
re-vaccination of children with naturally acquired immunity 
[@Christenson1994-fj]. With this draw parameters function, we incorporated 
titer-dependent boosting which will take into account an individual's current 
titer and adjust their realized boost accordingly. We arbitrarily set the 
titer-dependent boosting effects following an additional successful exposure 
event for individuals above 200 mIU/mL to be only 10% of the full boost.

Given all of our specified parameters and models, individuals in the 
simulation can only be vaccinated once and naturally infected once. If an 
individual in vaccinated and their titers are below the threshold of protection 
they run the risk of experiencing a natural infection where their titers will 
receive a partial boost. Given the short duration (10 years) of this simulation 
and the vaccine associated antibody kinetics parameters, it is unlikely that any 
individual who is vaccinated and successfully immunized will reach a titer level 
that puts them at risk of infection. If an individual is vaccinated following a 
natural infection event, then they will also receive a partial boost. 


```{r}
## Specify the antibody model 
antibody_model<-antibody_model_biphasic

## Bring in the antibody parameters needed for the antibody model
## Note that the titer-mediated protection parameters needed for the immunity model (Section 1.5), the titer-ceiling parameters needed for draw_parameters and the observation error parameter needed for the observation model (Section 1.7) are all defined here too.
model_pars_path <- system.file("extdata", "model_pars_cs1.csv", package = "serosim")
model_pars_original <- read.csv(file = model_pars_path, header = TRUE)
model_pars_original 

## Reformat model_pars for runserosim
model_pars<-reformat_biomarker_map(model_pars_original)
model_pars

## Specify the draw_parameters function to use 
draw_parameters<-draw_parameters_random_fx_biomarker_dep

## Plot biomarker dependent boosting effects given parameters specified within model_pars for biomarker 1 (measles_IgG)
plot_biomarker_dependent_boosting(start=0, end=1000, by=1, biomarker_ceiling_threshold=200, biomarker_ceiling_gradient=.0045)

## These parameters indicate that any individual with a titer level above 200 mIU/mL at the time of a second exposure event will only receive 10% of the regular boosting parameter for that event. Individuals with a measles titer of 0 will receive the full boost.
```

```{r, fig.dim = c(5, 4)}
## Plot example biomarker trajectories given the specified antibody kinetics model, model parameters and draw parameters function 
plot_antibody_model(antibody_model_biphasic, N=100, model_pars=model_pars,draw_parameters_fn = draw_parameters_random_fx_biomarker_dep, biomarker_map=biomarker_map)
```

# 1.7 Observation Model and observation_times
Now we specify how observed biomarker quantities, in this case antibody titers, are generated as a probabilistic 
function of the true, latent biomarker quantities, and when to observe these quantities.
In this step, users are specifying the sampling design and assay choice for 
their serological survey. 

We will take paired samples of all individuals alive at the midpoint (t=60) and 
end of the simulation (t=120). Our selected observation model observes the 
latent titer values given a continuous assay with user-specified lower and upper 
limits and added noise. Enzygnost ELISA measles IgG kits' lower limit of 
detection is 100 mIU/mL with no reported upper limit. The added noise represents 
assay variability and is implemented by sampling from a distribution with the 
true latent antibody titer as the mean and the measurement error as the standard 
deviation. The observation standard deviation and distribution are defined 
within model_pars as the “obs_sd” parameter. Within this observation model, we 
can also specify the assay sensitivity and specificity. The Enzygnost ELISA IgG 
test kit has a reported 99.6% sensitivity and a 100% specificity as reported by 
the manufacturer. 
```{r}
## Specify the limits of detection for each biomarker for the continuous assays
bounds<-tibble(biomarker_id=c(1,1),name=c("lower_bound","upper_bound"),value=c(100,Inf))

## Specify the observation model 
observation_model<-observation_model_continuous_bounded_noise

## Specify assay sensitivity and specificity needed for the observation model
sensitivity<-0.996
specificity<-1

## Specify observation_times (serological survey sampling design) to observe biomarker 1 (measles IgG antibody titer) across all individuals at the midpoint and the end of the simulation (t=60 and t=120)
obs1 <- tibble(i=1:N,t=60, b=1)
obs2 <- tibble(i=1:N,t=120, b=1)
observation_times<-rbind(obs1,obs2)
```


# 1.8 Optional Arguments 
Users can specify any additional arguments needed for their models. Here, we 
specify an optional argument to print an message to update us on the progress of 
the simulation. 
```{r}
## If VERBOSE is specified within runserosim, there will be an update message printed for every multiple of the integer specified.
## Print a message for every 10 individuals 
VERBOSE<-10
```

# 1.9 Run Simulation 
This is the core simulation where all simulation settings, models and parameters 
are specified within the main simulation function. The run time for this step varies 
depending on the number of individuals and the complexities of the specified models. 
There is a built in pre-computation step within _runserosim_ where the simulation attempts to perform as much pre-computation as possible for the exposure model to speed up the main simulation code. Users can turn off this pre-computation by setting pre-computation to FALSE within _runserosim_.
```{r message=FALSE, warning=FALSE, eval=TRUE}
## Run the core simulation and save outputs in "res"
res<- runserosim(
  simulation_settings,
  demography,
  observation_times,
  foe_pars,
  biomarker_map,
  model_pars,
  exposure_model,
  immunity_model,
  antibody_model,
  observation_model,
  draw_parameters,

  ## Specify other arguments needed
  VERBOSE=VERBOSE,
  bounds=bounds,
  max_events=max_events,
  vacc_exposures=vacc_exposures,
  vacc_age=vacc_age,
  sensitivity=sensitivity,
  specificity=specificity
)

## Note that models and arguments specified earlier in the code can be specified directly within this function.
```

# 1.10 Generate Plots
Now that the simulation is complete, let's plot and examine the simulation outputs.
```{r, fig.dim = c(6, 4)}
## Plot biomarker kinetics and exposure histories for 10 individuals 
plot_subset_individuals_history(res$biomarker_states, res$exposure_histories_long, subset=10, demography)
```

```{r, fig.dim = c(4, 6)}
## Plot individual probability of exposure for all exposure types.
## This is the output of the exposure model.
## Note: All individuals are under the same force of exposure since we specified a simple exposure model and constant foe_pars
plot_exposure_force(res$exposure_force_long)
```

```{r,fig.dim = c(4, 6)}
## Plot individual successful exposure probabilities for all exposure types
## This is the output of the exposure model multiplied by the output of the immunity model.
## In other words, this is the probability of exposure event being successful and inducing an immunological response
plot_exposure_prob(res$exposure_probabilities_long)
```

```{r,fig.dim = c(4, 6)}
## Plot individual exposure histories for all exposure types
plot_exposure_histories(res$exposure_histories_long)
```

```{r, fig.dim = c(4, 6)}
## Plot true biomarker quantities for all individuals across the entire simulation period
plot_biomarker_quantity(res$biomarker_states)
```

```{r,fig.dim = c(4, 5)}
## Plot the serosurvey results (observed biomarker quantities) at time 60 
obs60<-res$observed_biomarker_states %>% filter(t==60)
plot_obs_biomarkers_one_sample(obs60)
```

```{r, fig.dim = c(4, 5)}
## Plot the serosurvey results (observed biomarker quantities) at time 120 
obs120<-res$observed_biomarker_states %>% filter(t==120)
plot_obs_biomarkers_one_sample(obs120)
```

```{r,fig.dim = c(5, 5)}
## Plot both serosurveys paired samples
plot_obs_biomarkers_paired_sample(res$observed_biomarker_states)

## Note that the simulated kinetics parameters are also stored
head(res$kinetics_parameters)

```
