---
title: 'Paper case study 2: Multi-pathogen system with multivalent vaccinaton (Diphtheria and Pertussis)'
author: "Arthur Menezes"
date: '2022-08-24'
output: html_document
bibliography: cs2_DP_PT_citations.bib
link-citations: yes
---
Here, we will use the serosim package to generate a cross-sectional serosurvey of 
a multi-pathogen system with multivalent vaccination. The overall structure 
and principles discussed within this case study are applicable to pathogen 
systems like measles, mumps, and rubella with MMR vaccination or diphtheria, 
tetanus, and pertussis system with DTP vaccination among many others.  This 
simulation will track natural infection to pathogen A, natural infection to 
pathogen B and bi-valent vaccination with pathogen A and pathogen B antigens for 
100 individuals across a 10 year period. At the end of the 10 year period, we 
will measure titers against both pathogen A specific antibodies and pathogen B 
specific antibodies for all individuals. We will set up each of the required 
arguments and models for runserosim in the order outlined in the methods section 
of the paper. 

This case study is built around diphtheria and pertussis and we are interested 
in tracking bivalent vaccination and both diphtheria and pertussis natural infection 
across a 10 year period. We will conduct a cross-sectional serological survey 
where we will use an ELISA kit to measure an individual's diphtheria and 
pertussis IgG titer.

We will use published diphtheria and pertussis serological surveys and any well 
characterized epidemiological and immunological characteristics of diphtheria 
and pertussis to parameterize our models wherever possible and make reasonable 
assumptions for other required model parameters wherever information is lacking. 

We narrowed our search to diphtheria and pertussis serological surveys conducted 
with ELISA IgG test kits that reported quantitative titers in comparable units. 
We did not conduct an extensive search of all diphtheria and pertussis serological surveys in 
order to parameterize our models as it is beyond the scope of this case study. 
Our aim was to provide a simple example of how users can use the literature to 
inform required serosim inputs. Within each section below, we will briefly 
explain the rationale behind our selected model inputs. We caution users to 
conduct their own research into the models and associated parameters which best 
align with their disease system and biomarker test kits. 


Load necessary packages:
```{r message=FALSE, warning=FALSE,eval=TRUE}
devtools::load_all("~/Documents/GitHub/serosim")
library(tidyverse)
library(data.table)
library(ggplot2)
library(patchwork)
library(reshape2)
```


# 1.1 Simulation Settings
We will simulate monthly time steps across a 10 year period. Therefore, we will 
have 120 time steps. Note that these are arbitrary time steps which will need to 
be scaled to the right time resolution to match any time-based parameters used 
later on.
```{r}
## Specify the number of time periods to simulate 
times <- seq(1,120,by=1) 

## Set simulation settings
simulation_settings <- list("t_start"=1,"t_end"=max(times))
```

# 1.2 Population Demography
For this case study, we are interested in tracking an individual's socioeconomic 
status, group(a proxy for location), birth and removal time. We will use the 
generate_pop_demography function to create the demography tibble needed for 
runserosim.

Note: The runserosim function, called later on, only requires a demography 
tibble with two columns (individuals and times). 

```{r}
## Specify the number of individuals in the simulation 
N<-100

## Pre load the demography categories, values and distributions 
## Specify options for each demography element and the distribution of each within the population
## We are interested in simulating a population where individuals have varying nutritional status and can reside in either of 2 locations
aux <- list("NS"=list("name"="NS","options"=c("low","medium","high"), "distribution"=c(0.3,0.3,0.4)),
            "Group"=list("name"="group","options"=c("1", "2"), "distribution"=c(0.5,0.5)))


## Generate the population demography tibble
## Let's assume that individuals are removed from the population and set prob_removal to 0.2
demography <- generate_pop_demography(N, times, age_min=0, removal_min=1, removal_max=120, prob_removal=0.2, aux=aux)

## Examine the generated demography tibble
head(demography)
tail(demography)
```

# 1.3 Exposure to biomarker mapping
Now, we have to set up the exposure IDs and biomarker IDs for our desired 
simulation. Individuals can be seropositive either by natural infection or by 
vaccination and we want to track both exposure types separately. 

In this case, we need three exposure IDs:

1. Diphtheria natural infection (DP_ifxn)
2. Pertussis natural infection (PT_ifxn)
3. Diphtheria and pertussis combined vaccination (vacc)

And two biomarker IDs:

1. Diphtheria specific antibodies (DP_antibody)
2. Pertussis specific antibodies (PT_antibody)

runserosim requires that exposure_id and biomarker_id are numeric so we will use 
the reformat_biomarker_map function to create a new version of the biomarker map. 
Users can go directly to numeric biomarker_map if they wish. Note that the 
reformat_biomarker_map function  will number the exposures and biomarkers in 
alphabetical order so that the first exposure event or biomarker that is listed
will not necessarily be labeled as 1.
```{r}
## Create biomarker map
biomarker_map_original <- tibble(exposure_id=c("DP_ifxn","PT_ifxn","vacc","vacc"),biomarker_id=c("DP_antibody","PT_antibody","DP_antibody","PT_antibody"))
biomarker_map_original

## Reformat biomarker_map for runserosim
biomarker_map <-reformat_biomarker_map(biomarker_map_original)
biomarker_map
```

# 1.4 Force of Exposure and Exposure Model
Now, we need to specify the foe_pars argument which contains the force of exposure
for all exposure_IDs across all time steps and groups. We also need to specify the 
exposure model which will be called within runserosim later. The exposure model 
will determine the probability that an individual is exposed to a specific 
exposure event.

We specified the same force of exposure for all time steps within foe_pars for 
simplicity but users will likely have varying numbers to match real world settings. 

For our selected exposure model, the probability of exposure (1-e^(-λ)) depends 
on the force of exposure (λ) at the current time t for group g modulated by 
relevant demographic elements (dem_mod). For simplicity, we 
selected arbitrary age and demography modifiers specified below. Since we have 
two groups, there are 2 rows in the first foe_pars dimension.


```{r}
## Create an empty array to store the force of exposure for all exposure types across all time steps and groups
## Dimension 1: Group
## Dimension 2: Times
## Dimension 3: Exposure ID in the numeric order that they appear in the biomarker map
foe_pars <- array(0, dim=c(n_distinct(demography$group),max(times), n_distinct(biomarker_map$exposure_id)))

## Note that we can specify a different force of exposure for each group, time and exposure ID
## We specified the same value for all time steps within foe_pars for simplicity.

## Specify the force of exposure for exposure ID 1 which represents diphtheria natural infection (DP_ifxn)
foe_pars[1,,1] <- 0.04 ## Group 1 (aka Location 1)
foe_pars[2,,1] <- 0.03 ## Group 2 (aka Location 2)

## Specify the force of exposure for exposure ID 2 which represents pertussis natural infection (PT_ifxn)
foe_pars[1,,3] <- 0.02 ## Group 1 (aka Location 1)
foe_pars[2,,3] <- 0.01 ## Group 2 (aka Location 2)

## Specify the force of exposure for exposure ID 3 which represents diphtheria and pertussis combined vaccine (vacc)
foe_pars[1,,2] <- 0.02 ## Group 1 (aka Location 1)
foe_pars[2,,2] <- 0.03 ## Group 2 (aka Location 2)

## Specify a simple exposure model which calculates the probability of exposure from the force of exposure modulated by age and demography elements 
exposure_model<-exposure_model_dem_mod

## This exposure model requires dem_mod and t_in_year arguments


## Create a tibble with any relevant demography modifiers that affect exposure probability 
## For simplicity, we selected arbitrary numbers. 

## First, we will specify age modifiers 
## Individuals who are 0-3 are 2 times more likely to be exposed to diphtheria
age_mod_1<-tibble(exposure_id=rep(1,11), column=rep("age",times=11), value=0:10, modifier=c(2,2,2,2,1,1,1,1,1,1,1))

## Individuals who are 0-3 are 2 times more likely to be exposed to pertussis
age_mod_2<-tibble(exposure_id=rep(2,11), column=rep("age",times=11), value=0:10, modifier=c(2,2,2,1,1,1,1,1,1,1,1))

## Individuals who are less than one year old are 3 times more likely be vaccinated than other age classes
age_mod_3<-tibble(exposure_id=rep(3,11), column=rep("age",times=11),  value=0:10, modifier=c(3,1,1,1,1,1,1,1,1,1,1))
age_mod<-rbind(age_mod_1,age_mod_2,age_mod_3)
age_mod

## Now we will  specify additional demography exposure modifiers and combine them with the previous ones
## Here, individuals who are of low nutritional status are twice as likely of being exposed to diphtheria and pertussis while individuals who are of medium nutritional status are 1.5 times as likely of being exposed when compared to individuals of high nutritional status 
## Individuals of high nutritional status are 3 times more likely to be exposed to exposure ID 3 (vaccination) while individuals who are of medium nutritional status are 2 times more likely to be exposed to exposure ID 3 (vaccination) than individuals of low nutritional status. 
## Note that the modifiers must be defined for all combinations of exposure types and demographic elements
mod<-tibble(exposure_id=c(1,1,1,2,2,2,3,3,3), column=rep("NS",times=9), value=rep(c("low","medium", "high"),3), modifier=c(2,1.5,1,2,1.5,1,1,2,3))
mod
## Combine both age modifiers and additional modifiers
dem_mod<-rbind(age_mod,mod)

## Specify the number of time steps within a year 
## We are simulating on the monthly scale
t_in_year=12
```

# 1.5 Immunity Model
Here, we specify the immunity model which will determine the probability that an 
exposure event is successful. Since we have both vaccination and natural 
infection events, we will use immunity_model_vacc_ifxn_titer_prot. With this 
immunity model, the probability of successful vaccination exposure depends on 
the number of vaccines received prior to time t and the individual's age at time 
t while the probability of successful infection is dependent on the titer at the 
time of exposure. Individuals can have up to 3 successful vaccination 
events (aka 3 doses) and are eligible for vaccination starting at 2 months of age 
to align with the diphtheria/pertussis vaccination schedule. 

As for the probability of a successful infection, we placed no limit on the 
maximum number of infections to either diphtheria or pertussis however our 
selected immunity model will take into account an individual's current titer 
level when determining the probability of a successful infection. Additional 
successful exposure events aka "reinfection" events are representative of 
boosting events which can occur if an individual is exposed to either pathogen 
following a vaccination or previous infection. The user can limit the frequency 
of  reinfections by adjusting the titer-mediated protection parameters or by 
specifying a maximum number of allowed successful exposure events The 
titer-mediated protection parameters used within this model are defined within 
model_pars which will be loaded in section 1.6. 

Again, the following diphtheria and pertussis titer-mediated protection parameters were selected by a non-exhaustive literature search. The titer cutoffs used below were also discussed in a previous 
analysis of diphtheria and pertussis serology [@Razafimahatratra2022-ub].

Diphtheria:
1. Individuals whose diphtheria antibody levels are under 0.01 IU/mL are considered highly susceptible to disease while individuals with higher levels are associated with less severe symptoms [@Ipsen1946-qc; @Bjorkholm1986-qk; @Danilova2006-vj; @Ohuabunwo2005-rn]. 
2. Individuals whose antibody levels are 0.01 IU/mL have the lowest titer level which confers some degree of protection while individuals with >=0.1 IU/mL are associated with long term protection [@Efstratiou1994-xh,].
3. Antibody levels in between 0.01 and 0.09 IU/mL are considered to provide basic levels of protection against disease [@World_Health_Organization2017-hc].

Pertussis:
1. Low anti-pertussis IgG antibodies have shown significant correlation with pertussis susceptibility [@Storsaeter2003-uk] but a protective threshold has not been established [@Saso2020-gd] since there is no established immunological correlate of protection against disease [@Plotkin2008-oi]
2. Given the information presented above, we have chosen that antibody levels between 40-100 IU/mL will confer higher level of protection since they are indicative of serpositivity and of recent pertussis infection or pertussis vaccination [@noauthor_undated-hp; @Chen2016-ka].

```{r}
## Specify immunity model within the runserosim function below 
immunity_model<-immunity_model_vacc_ifxn_titer_prot

## Specify which exposure IDs represent vaccination events 
## In this case, only exposure ID 3 is a vaccination event
vacc_exposures<-3

## Specify the time step after birth at which an individual is eligible for vaccination (2 months old for diphtheria and pertussis combine vaccine); ; note non vaccine exposures are listed as NAs
vacc_age<-c(NA,NA,2)

## Specify the maximum number of successful exposure events an individual can receive for each exposure type
## We placed no successful exposure limit on the number of diphtheria or pertussis infection exposures and 3 dose limit on the vaccine exposure.
max_events<-c(Inf,Inf,3)

## Plot titer-mediated protection curve given parameters specified within model_pars for biomarker 1, diphtheria antibody  (DP_antibody) which will be loaded in section 1.6
## The immunity model we selected will take into account an individual's current titer to diphtheria when determining the probability of successful infection. 
## Titers are ploted in mIU/mL 
plot_titer_mediated_protection(0:150, titer_prot_midpoint=75, titer_prot_width=.1)

## Plot titer-mediated protection curve given parameters specified within model_pars for biomarker 2, pertussis antibody (PT_antibody) which will be loaded in section 1.6
## The immunity model we selected will take into account an individual's current titer to pertussis when determining the probability of successful infection. 
## Titers are ploted in IU/mL 
plot_titer_mediated_protection(0:100, titer_prot_midpoint=40, titer_prot_width=.25)
```

# 1.6  Antibody model and model parameters 
Now, we specify the antibody model to be used within runserosim. We will be 
using a biphasic boosting-waning model. This model assumes that for each 
exposure there is a set of long-term boost, long-term boost waning, short-term 
boost, and short-term boost waning parameters. Given varying diphtheria 
antibody dynamics observed in early and late life, we decided that a biphasic 
boosting-waning model was appropriate for our purposes
[@Voysey2016-gu].

The antibody kinetics parameters needed for the antibody model are pre-loaded
within a csv file which will take on the argument name: model_pars.
Users can edit the csv file to modify any parameters. runserosim requires that 
exposure_id and biomarker_id are numeric within model_pars so we will use the 
reformat_biomarker_map function again to create a new version of model_pars. 
Users can go directly to numeric model_pars if they wish. 

We selected a few diphtheria and pertussis serological surveys conducted which 
reported IgG titers following vaccination to inform the vaccine induced boost 
and waning parameters. There are many complexities and heterogeneity in these 
parameters which arise from different vaccine types and vaccination schedules. 
We neglect these complexities and select a few papers to give us some rough 
parameters for this example simulation.

After 3 vaccine doses, diphtheria titers ranged 1.5-1.7 IU/mL [@Kimura1991-pg] and 
pertussis titers above 100 IU/mL are associated with recent vaccination 
[@noauthor_undated-hp]. Diphtheria antibody half-life has been estimated to be around 
19-27 years [@Amanna2007-pf; @Hammarlund2016-lj] with 10% of children losing immunity by 
one year following the primary vaccination series [@Pichichero1987-sl]; 
67% of children after 3 to 13 years and 83% after 14 to 23 years 
[@Crossley1979-ta]. As for pertussis, the duration of immunity after a 
3 dose series of the wP vaccine is estimated to be from 4 to 12 years 
[@Sheridan2014-fo]. Pertussis titers >=100 IU/mL are associated with recent 
infection or vaccination within the past year while titers between 40 and 100IU/mL 
are associated with recent infection [@noauthor_undated-hp]

Our selected antibody kinetics parameters given our brief and limited literature 
search are as follows:

Diphtheria vaccine:
Long term boost: 0.3-0.5 IU/mL
Long term boost waning: 0.01 IU/mL per month
Short term boost: 0.1-0.3 IU/mL
Short term boost waning: 0.033 IU/mL per month

Pertussis vaccine:
Long term boost: 30-50 IU/mL
Long term boost waning: 0.0066 IU/mL per month
Short term boost: 40-80 IU/mL
Short term boost waning: 0.016 IU/mL per month

We were unable to find studies that reported titer kinetics following known 
infection events in comparable unit. For simplicity, we will assume that natural 
infection boosting parameters are 25% higher than vaccine induced antibody 
boosting with similar waning rates.

Lastly, we define the draw_parameters function which determines how each 
individual’s antibody kinetics parameters are simulated from the within-host 
processes parameters tibble (model_pars). We will use a function which draws 
parameters directly from model_pars for the antibody model with random 
effects to represent individual heterogeneity in immunological responses. 
Parameters are drawn randomly from a distribution with mean and standard 
deviation specified within model_pars. This draw_parameters function also 
incorporates titer-dependent boosting where an individual’s realized boost is 
dependent on their titer level at the time of the exposure event.

We arbitrarily set the diphtheria titer-dependent boosting effects following an 
additional successful exposure event for individuals with diphtheria levels above 
1.7 to be only 10% of the full boost. Similarly, individuals with pertussis 
titers above 100 IU/mL will only receive 10% of the full boost for any 
additional successful exposure event. 
```{r}
## Specify the antibody model 
antibody_model<-antibody_model_biphasic

## Bring in the antibody parameters needed for the antibody model 
## Note that the titer-mediated protection parameters needed for the immunity model, the titer-ceiling parameters needed for draw_parameters and the observation error parameter needed for the observation model are all defined here too.
## Also note that these are all arbitrary parameter values chosen for this toy example.
model_pars_path <- system.file("extdata", "model_pars_cs2.csv", package = "serosim")
model_pars_original <- read.csv(file = model_pars_path, header = TRUE)
model_pars_original 

## Reformat model_pars for runserosim
model_pars<-reformat_biomarker_map(model_pars_original)
model_pars

## Specify the draw_parameters function to use 
draw_parameters<-draw_parameters_random_fx_titer_dep

## Plot titer dependent boosting effects given parameters specified within model_pars for biomarker 1, diphtheria antibody (DP_antibody)
plot_titer_dependent_boosting(start=0, end=2, by=.1, titer_ceiling_threshold=1.7, titer_ceiling_gradient=0.529411)

## Plot titer dependent boosting effects given parameters specified within model_pars for biomarker 2, pertussis antibody (PT_antibody)
plot_titer_dependent_boosting(start=0, end=125, by=1, titer_ceiling_threshold=100, titer_ceiling_gradient=0.009)

```

# 1.7 Observation Model 
Now we specify how observed antibody titers are generated as a probabilistic 
function of the true, latent antibody titer and when to observe these titers. 
In this step, users are specifying the sampling design and assay choice for 
their serological survey. 

This observation model observes the latent titer values 
given a continuous assay with user-specified lower and upper limits, added 
noise. The added noise represents assay variability and is done by sampling 
from a distribution with the latent antibody titer as the mean and the measurement 
error as the standard deviation. The observation standard deviation and distribution 
is defined within model_pars as the “obs_sd” parameter for each biomarker.

Diphtheria:
Lower limit of ELISA IgG Detectability: 0.01 IU/mL
Upper limit of ELISA IgG Detectability: 2 IU/mL

Pertussis:
Lower limit of ELISA IgG Detectability: 5 IU/mL
Upper limit of ELISA IgG Detectability: 200 IU/mL


```{r}
## Specify the limits of detection for each biomarker for the continuous assays
bounds<-tibble(biomarker_id=c(1,1,2,2),name=rep(c("lower_bound","upper_bound"),2),value=c(0.01,2,5,200))

## Specify the observation model 
observation_model<-observation_model_continuous_bounded_noise

## Specify observation_times to observe both biomarkers (aka DP_antibody and PT_antibody titers) across all individuals at the end of the simulation (t=120)
obs1 <- tibble(i=1:N,t=120, b=1)
obs2 <- tibble(i=1:N,t=120, b=2)
observation_times<-rbind(obs1,obs2)
```


# 1.8 Optional Arguments 
There are no optional arguments needed for this simulation. 

# 1.9 Run Simulation 
This is the core simulation where all simulation settings, models and parameters 
are specified within the main simulation function. The run time for this step varies 
depending on the number of individuals and the complexities of the specified models. 
```{r message=FALSE, warning=FALSE, eval=TRUE}
res<- runserosim(
  simulation_settings,
  demography,
  observation_times,
  foe_pars,
  biomarker_map,
  model_pars,
  exposure_model,
  immunity_model,
  antibody_model,
  observation_model,
  draw_parameters,

  ## Other arguments needed
  bounds=bounds,
  max_events=max_events,
  vacc_exposures=vacc_exposures,
  vacc_age=vacc_age,
  dem_mod=dem_mod,
  t_in_year=t_in_year
)

## Note that models and arguments specified earlier in the code can be specified directly within this function.
```

# 1.10 Generate Plots
Now that the simulation is complete, let's plot and examine the simulation outputs.
```{r} 
## Plot diphtheria antibody kinetics and exposure histories for 10 individuals 
plot_subset_individuals_history(res$antibody_states %>% filter(b==1), res$exposure_histories_long %>% filter(x==1 | x==3), subset=10, demography, removal=TRUE)

## Plot pertussis antibody kinetics and exposure histories for 10 individuals 
plot_subset_individuals_history(res$antibody_states %>% filter(b==2), res$exposure_histories_long %>% filter(x==2 | x==3), subset=10, demography, removal=TRUE)

## Plot individual probability of exposure for all exposure types.
## This is the output of the exposure model.
## Note: All individuals are under the same force of exposure since we specified a simple exposure model
plot_exposure_force(res$exposure_force_long)

## Plot individual successful exposure probabilities for all exposure types
## This is the output of the exposure model multiplied by the output of the immunity model.
## In other words, this is the probability of exposure event being successful and inducing an immunological response
plot_exposure_prob(res$exposure_probabilities_long)

## Plot individual exposure histories for all exposure types
plot_exposure_histories(res$exposure_histories_long)

## Plot diphtheria (biomarker 1) antibody states for all individuals
plot_titers(res$antibody_states %>% filter(b==1))

## Plot pertussis (biomarker 2) antibody states for all individuals
plot_titers(res$antibody_states %>% filter(b==2))

## Plot the diphtheria serosurvey results 
plot_obs_titers_one_sample(res$observed_antibody_states %>% filter(b==1))

## Plot the pertussis serosurvey results 
plot_obs_titers_one_sample(res$observed_antibody_states %>% filter(b==2))

## Note that the simulated kinetics parameters are also stored
head(res$kinetics_parameters)
```
